\textbf{CSDS 491: Assignment 4}

### 2021-04-22 A4/article.tex
✨ Add answeres to E4.

>>>>

Please refer to \ilc{A4/code/e4.py} for code implementation. Note the plotting is refered to \href{https://matplotlib.org/3.2.2/gallery/statistics/confidence_ellipse.html#sphx-glr-gallery-statistics-confidence-ellipse-py}{this}, and the GMM is refered to \href{http://ethen8181.github.io/machine-learning/clustering/GMM/GMM.html#Parameter-Estimation}{this} tutorials.

\subsection*{4.1}

\section{4.1}
The EM algorithm iteratively performs the \ilc{expectation} step and \ilc{maximization} step -- thus ``EM'' -- until the log-likelihood (the indicator of convergence) falls into an acceptable \ilc{thlr}.

\subsection*{Expectation Step}

The expectation step intends to solve the inference problem: which Gaussian distribution generated each of the data point?
By applying Baye's rule, the conditional probability of giving a data point $z_i = m$ given $z_i \in N_k$ can be represented as:
\begin{align*}
    \gamma_k = p(z_i=m|N_k) &= \frac{p(z_i=k)p(N_k|z_i=m)}{p(N_k)} \\
    &= \frac{p(z_i=k)p(N_k|z_i=m)}{\sum_{j=1}^Kp(z_i=j)p(N_k|z_i=j)} \\
    &= \frac{\pi_k\mathcal N(N_k|\mu_k, \Sigma_k)}{\sum_{j=1}^K \mathcal N(N_k|\mu_j, \Sigma_j)}
\end{align*}
Based on the parameters $\gamma_k$, $\pi_k$, $\mu_k$, and $\Sigma_k$, we are able to compute the log-likelihood for the purpose of optimization:
\begin{align*}
    \sum_k\sum_i \gamma_k^{(i)}log(\pi_k) + \sum_k\sum_i \gamma_{k}^{(i)}log(\mathcal N(N_k^{(i)|\mu_k, \Sigma_k}))
\end{align*}
The stopping criteria for the iteration is when the change of likelihood falls into a tolerance range that we predefined. We also need to define the
maximum number of iterations in order to prevent it runs into infinite loop.

\subsection*{Maximization Step}

Since the change of the membership of data points will also affect the distributions, we can derive a closed form updates for all the parameters (the means $\mu$, the responsibility $\pi$, and the covariance matrix $\Sigma$).
\begin{align*}
    \mu_k &= \frac{1}{N_k}\sum_{n=1}^{N}\gamma_k^{(n)}x^{(n)} \\
    \Sigma_{k} &= \frac{1}{N_k}\sum_{n=1}^{N}(x^{n}-\mu_{k})(x^{n}-\mu_{k})^T \\
    \pi_k &= \frac{N_k}{N}
\end{align*}
, where $N_k=\Sigma_{n=1}^{N}\gamma_{k}^{(n)}$. We will perform E-step and M-step interchangeably to optimize the classification of data points.

\subsection*{4.2}
\begin{figure}[H]
    \centering
    \includegraphics[width=0.5\linewidth]{{fig/e4.2}.png}
\end{figure}

\subsection*{4.3}

\begin{figure}[H]
\minipage{0.3\textwidth}
  \includegraphics[width=\linewidth]{{fig/e4.3_0}.jpg}
\endminipage\hfill
\minipage{0.3\textwidth}
  \includegraphics[width=\linewidth]{{fig/e4.3_1}.png}
\endminipage\hfill
\minipage{0.3\textwidth}
    \includegraphics[width=\linewidth]{{fig/e4.3_2}.png}
\endminipage
\end{figure}

We can tell that all species from the Iris dataset (on two attributes) are nicely clustered within their sigma contours.

\subsection*{4.4}

\subsubsection*{K = 1}

\begin{figure}[H]
\minipage{0.3\textwidth}
  \includegraphics[width=\linewidth]{{fig/e4.4_k1_0}.png}
\endminipage\hfill
\minipage{0.3\textwidth}
    \includegraphics[width=\linewidth]{{fig/e4.4_k1_5}.png}
\endminipage\hfill
\minipage{0.3\textwidth}
    \includegraphics[width=\linewidth]{{fig/e4.4_k1_10}.png}
\endminipage
\end{figure}

Where the $\mu, \Sigma, \pi$ for $K = 1$ are respectively:
\begin{lstlisting}
mean: [[ 3.48778309 70.89705882]]
cov: [[[  1.29793889  13.92641885]
  [ 13.92641885 184.14381488]]]
pi: [1.]

mean: [[ 3.48778309 70.89705882]]
cov: [[[  1.29793889  13.92641885]
  [ 13.92641885 184.14381488]]]
pi: [1.]

mean: [[ 3.48778309 70.89705882]]
cov: [[[  1.29793889  13.92641885]
  [ 13.92641885 184.14381488]]]
pi: [1.]
\end{lstlisting}

\subsubsection*{K = 2}

\begin{figure}[H]
\minipage{0.3\textwidth}
  \includegraphics[width=\linewidth]{{fig/e4.4_k2_0}.png}
\endminipage\hfill
\minipage{0.3\textwidth}
    \includegraphics[width=\linewidth]{{fig/e4.4_k2_5}.png}
\endminipage\hfill
\minipage{0.3\textwidth}
    \includegraphics[width=\linewidth]{{fig/e4.4_k2_10}.png}
\endminipage
\end{figure}

Where the $\mu, \Sigma, \pi$ for $K = 2$ are respectively:
\begin{lstlisting}

mean: [[ 4.32816932 81.43143106]
[ 2.71366246 61.19333632]]
cov: [[[  0.14492959   0.45361294]
  [  0.45361294  28.18429193]]

[[  1.11020876  10.67014897]
  [ 10.67014897 131.42085961]]]
pi: [0.47947807 0.52052193]

mean: [[ 4.32816932 81.43143106]
[ 2.71366246 61.19333632]]
cov: [[[  0.14492959   0.45361294]
  [  0.45361294  28.18429193]]
[[  1.11020876  10.67014897]
  [ 10.67014897 131.42085961]]]
pi: [0.47947807 0.52052193]

mean: [[ 4.32816932 81.43143106]
[ 2.71366246 61.19333632]]
cov: [[[  0.14492959   0.45361294]
  [  0.45361294  28.18429193]]

[[  1.11020876  10.67014897]
  [ 10.67014897 131.42085961]]]
pi: [0.47947807 0.52052193]
\end{lstlisting}

\subsubsection*{K = 3}

\begin{figure}[H]
\minipage{0.3\textwidth}
  \includegraphics[width=\linewidth]{{fig/e4.4_k3_0}.png}
\endminipage\hfill
\minipage{0.3\textwidth}
    \includegraphics[width=\linewidth]{{fig/e4.4_k3_5}.png}
\endminipage\hfill
\minipage{0.3\textwidth}
    \includegraphics[width=\linewidth]{{fig/e4.4_k3_10}.png}
\endminipage
\end{figure}
Where the $\mu, \Sigma, \pi$ for $K = 3$ are respectively:
\begin{lstlisting}

mean: [[ 4.23203926 83.42947492]
[ 2.08519156 55.01970556]
[ 4.34563196 78.61168712]]
cov: [[[ 0.1506054   0.52994426]
  [ 0.52994426 25.80940221]]
[[ 0.13317075  1.1482405 ]
  [ 1.1482405  41.28312871]]
[[ 0.1526638   1.03335337]
  [ 1.03335337 30.27117787]]]
pi: [0.20632691 0.36913676 0.42453633]

mean: [[ 4.23203926 83.42947492]
[ 2.08519156 55.01970556]
[ 4.34563196 78.61168712]]
cov: [[[ 0.1506054   0.52994426]
  [ 0.52994426 25.80940221]]
[[ 0.13317075  1.1482405 ]
  [ 1.1482405  41.28312871]]
[[ 0.1526638   1.03335337]
  [ 1.03335337 30.27117787]]]
pi: [0.20632691 0.36913676 0.42453633]

mean: [[ 4.23203926 83.42947492]
[ 2.08519156 55.01970556]
[ 4.34563196 78.61168712]]
cov: [[[ 0.1506054   0.52994426]
  [ 0.52994426 25.80940221]]
[[ 0.13317075  1.1482405 ]
  [ 1.1482405  41.28312871]]
[[ 0.1526638   1.03335337]
  [ 1.03335337 30.27117787]]]
pi: [0.20632691 0.36913676 0.42453633]
\end{lstlisting}
